Water, High-Altitude Condensates, and Methane Depletion in the Atmosphere of the Warm Neptune WASP-107\lowercase{b}

### 2017-08-22 ms.tex
final version?

\begin{abstract}
The Neptune-mass exoplanet WASP-107b is an exciting target for atmosphere characterization. It has an unusually large atmospheric scale height and a small, bright host star, raising the possibility of precise constraints on its current nature and formation history.  We report the first atmospheric study of WASP-107b, a Hubble Space Telescope measurement of its near-infrared transmission spectrum.  We determined the planet's composition with two techniques: atmospheric retrieval based on the transmission spectrum and interior structure modeling based on the observed mass and radius. We detect water at $6.5\,\sigma$ confidence and set an upper limit on the atmospheric metallicity of $30\times$ solar. We find that the methane abundance is depleted relative to expectations for a solar abundance pattern (at $3\,\sigma$ confidence), suggesting a low carbon-to-oxygen ratio or high internal heat flux.  The water features are smaller than expected for a cloudless atmosphere, crossing less than one scale height, and require a thick condensate layer at high altitudes (0.1 - 3 mbar). We find that it is challenging for physically motivated cloud and haze models to produce opaque condensates at these pressures.  Taken together, these findings serve as an illustration of the diversity and complexity of exoplanet atmospheres. The community can look forward to more such results with the high precision and wide spectral coverage afforded by future observing facilities. 
\end{abstract}

%% Keywords should appear after the \end{abstract} command. 
%% See the online documentation for the full list of available subject
%% keywords and the rules for their use.
\keywords{planets and satellites: individual (WASP-107b), planets and satellites: atmospheres}

%% From the front matter, we move on to the body of the paper.
%% Sections are demarcated by \section and \subsection, respectively.
%% Observe the use of the LaTeX \label
%% command after the \subsection to give a symbolic KEY to the
%% subsection for cross-referencing in a \ref command.
%% You can use LaTeX's \ref and \label commands to keep track of
%% cross-references to sections, equations, tables, and figures.
%% That way, if you change the order of any elements, LaTeX will
%% automatically renumber them.

%% We recommend that authors also use the natbib \citep
%% and \citet commands to identify citations.  The citations are
%% tied to the reference list via symbolic KEYs. The KEY corresponds
%% to the KEY in the \bibitem in the reference list below.

@@

\section{Observations and Data Reduction}
We observed a single transit of WASP-107b with \emph{HST}'s Wide Field Camera 3 (WFC3) instrument on UT 5-6 June 2017.  The transit observation consisted of five \emph{HST} orbits. At the beginning of each 96-minute orbit, we took an image of the target with the F130N filter (exposure time = 4.2 s). This direct image is used for wavelength calibration. For the remainder of the target visibility period (about 45 minutes), we obtained time series spectra with the G141 grism, which provides low-resolution spectroscopy over the wavelength range $1.1 - 1.7\,\mu$m.  We used the NSAMP=6, SPARS\_25 readout mode (exposure time = 112 s) to optimize the efficiency of the observations, as determined by the \texttt{PandExo\_HST} planning tool \citep{batalha17}.  As is standard for WFC3 observations of bright targets, we used the spatial scanning observing mode, which slews the telescope in the spatial direction over the course of an exposure. The scan rate was 0.12 arcseconds/sec.

@@

\subsection{Spectroscopic Light Curve Fits}
We binned the spectrum into 20 spectrophotometric channels from 1.12 to 1.65 $\mu$m, shown in Figure\,\ref{fig:lc}. We fit the light curves with the \texttt{divide-white} technique, which assumes that the light curve systematics have the same morphology at all wavelengths \citep{stevenson14c, kreidberg14a}. For this method, the transit model $T_\lambda(t)$ is multiplied by the systematics vector from the white light curve fit ($F_\mathrm{white}/T_\mathrm{white}$), and rescaled by a factor $C_\lambda + V_\lambda t$.  One advantage of this approach is that it removes the star spot crossing feature, enabling us to use orbit three with no additional correction. The amplitude of the feature has no detectable wavelength dependence at the level of precision of our data.  As for the white light curve, we fixed some of the transit parameters on the estimates from \cite{dai17} and fixed the limb darkening on the PHOENIX model. The final spectroscopic light curve fits had just three free parameters per channel: $C_\lambda$, $V_\lambda$, and $r_p/r_s$.

@@

\begin{deluxetable}{c c c}
\tablecaption{WASP-107b transmission spectrum \label{tab:transit_depths}}
\tablehead{\colhead{Wavelength ($\mu$m)}  & \colhead{Transit depth} & \colhead{Uncertainty}}
\startdata
1.133 & 0.020641 & 5.9e-05 \\
1.158 & 0.020733 & 5.5e-05 \\
1.184 & 0.020505 & 5.6e-05 \\
1.209 & 0.020498 & 5.4e-05 \\
1.235 & 0.020455 & 5.9e-05 \\
1.260 & 0.020492 & 5.0e-05 \\
1.285 & 0.020620 & 6.2e-05 \\
1.311 & 0.020739 & 5.0e-05 \\
1.336 & 0.020660 & 5.7e-05 \\
1.362 & 0.020858 & 4.8e-05 \\
1.387 & 0.020794 & 4.8e-05 \\
1.413 & 0.020888 & 5.2e-05 \\
1.438 & 0.020821 & 6.2e-05 \\
1.464 & 0.020691 & 5.1e-05 \\
1.489 & 0.020682 & 6.9e-05 \\
1.515 & 0.020679 & 6.7e-05 \\
1.540 & 0.020509 & 6.0e-05 \\
1.565 & 0.020480 & 6.4e-05 \\
1.591 & 0.020500 & 5.6e-05 \\
1.616 & 0.020514 & 6.5e-05 \\
\enddata
\end{deluxetable}

@@

We also performed a ``free" retrieval that allowed the abundances of CH$_4$, H$_2$O and NH$_3$ to vary with no assumption of chemical equilibrium. The $3\,\sigma$ upper limit on methane abundance is $1.2\times10^{-3}$, ruling out the the predicted methane content for a composition with solar C/O and $3\,\times$ solar metallicity (the median metallicity from the chemical equilibrium retrieval).  Based on these tests, we conclude that the atmosphere of WASP-107b is depleted in methane relative to expectations for a solar abundance pattern. By contrast, the water abundance ($6\times10^{-6} - 2\times10^{-3}$) is consistent with predictions for solar composition.

>>>>

\begin{abstract}
The Neptune-mass exoplanet WASP-107b is an exciting target for atmosphere characterization. It has an unusually large atmospheric scale height and a small, bright host star, raising the possibility of precise constraints on its current nature and formation history.  We report the first atmospheric study of WASP-107b, a Hubble Space Telescope measurement of its near-infrared transmission spectrum.  We determined the planet's composition with two techniques: atmospheric retrieval based on the transmission spectrum and interior structure modeling based on the observed mass and radius. The interior structure models set a $3\,\sigma$ upper limit on the atmospheric metallicity of $30\times$ solar. The transmission spectrum shows strong evidence for water absorption ($6.5\,\sigma$ confidence), and we infer a water abundance consistent with expectations for a solar abundance pattern. On the other hand, methane is depleted relative to expectations (at $3\,\sigma$ confidence), suggesting a low carbon-to-oxygen ratio or high internal heat flux.  The water features are smaller than predicted for a cloudless atmosphere, crossing less than one scale height. A thick condensate layer at high altitudes (0.1 - 3 mbar) is needed to match the observations; however, we find that it is challenging for physically motivated cloud and haze models to produce opaque condensates at these pressures.  Taken together, these findings serve as an illustration of the diversity and complexity of exoplanet atmospheres. The community can look forward to more such results with the high precision and wide spectral coverage afforded by future observing facilities. 
\end{abstract}

\keywords{planets and satellites: individual (WASP-107b), planets and satellites: atmospheres}

@@

\section{Observations and Data Reduction}
We observed a single transit of WASP-107b with \emph{HST}'s Wide Field Camera 3 (WFC3) instrument on UT 5-6 June 2017.  The transit observation consisted of five \emph{HST} orbits. At the beginning of each 96-minute orbit, we took an image of the target with the F130N filter (exposure time = 4.2 s). This direct image is used for wavelength calibration. For the remainder of the target visibility period, we obtained time series spectra with the G141 grism, which provides low-resolution spectroscopy over the wavelength range $1.1 - 1.7\,\mu$m.  We used the NSAMP=6, SPARS\_25 readout mode (exposure time = 112 s) to optimize the efficiency of the observations, as determined by the \texttt{PandExo\_HST} planning tool \citep{batalha17}.  As is standard for WFC3 observations of bright targets, we used the spatial scanning observing mode, which slews the telescope in the spatial direction over the course of an exposure. The scan rate was 0.12 arcseconds/sec.

@@

\subsection{Spectroscopic Light Curve Fits}
We binned the spectrum into 20 spectrophotometric channels from 1.12 to 1.63 $\mu$m, shown in Figure\,\ref{fig:lc}. We fit the light curves with the \texttt{divide-white} technique, which assumes that the light curve systematics have the same morphology at all wavelengths \citep{stevenson14c, kreidberg14a}. For this method, the transit model $T_\lambda(t)$ is multiplied by the systematics vector from the white light curve fit ($F_\mathrm{white}/T_\mathrm{white}$), and rescaled by a factor $C_\lambda + V_\lambda t$.  One advantage of this approach is that it removes the star spot crossing feature, enabling us to use orbit three with no additional correction. The amplitude of the feature has no detectable wavelength dependence at the level of precision of our data.  As for the white light curve, we fixed some of the transit parameters on the estimates from \cite{dai17} and fixed the limb darkening on the PHOENIX model (coefficients listed in Table\,\ref{tab:transit_depths}. The final spectroscopic light curve fits had just three free parameters per channel: $C_\lambda$, $V_\lambda$, and $r_p/r_s$.

@@

\begin{deluxetable}{c c c c c}
\tablecaption{WASP-107b transmission spectrum and limb darkening coefficients\label{tab:transit_depths}}
\tablehead{\colhead{Wavelength ($\mu$m)}  & \colhead{Transit depth} & \colhead{Uncertainty} & \colhead{$u_1$} & \colhead{$u_2$}}
\startdata
1.121 -- 1.145 & 0.020641 & 5.9e-05 & 0.38 & 0.13 \\
1.145 -- 1.171 & 0.020733 & 5.5e-05 & 0.37 & 0.14 \\
1.171 -- 1.196 & 0.020505 & 5.6e-05 & 0.36 & 0.14 \\
1.196 -- 1.222 & 0.020498 & 5.4e-05 & 0.36 & 0.15 \\
1.222 -- 1.248 & 0.020455 & 5.9e-05 & 0.36 & 0.15 \\
1.248 -- 1.272 & 0.020492 & 5.0e-05 & 0.35 & 0.16 \\
1.272 -- 1.298 & 0.020620 & 6.2e-05 & 0.34 & 0.17 \\
1.298 -- 1.323 & 0.020739 & 5.0e-05 & 0.34 & 0.17 \\
1.323 -- 1.349 & 0.020660 & 5.7e-05 & 0.33 & 0.18 \\
1.349 -- 1.374 & 0.020858 & 4.8e-05 & 0.32 & 0.19 \\
1.374 -- 1.401 & 0.020794 & 4.8e-05 & 0.31 & 0.20 \\
1.401 -- 1.425 & 0.020888 & 5.2e-05 & 0.30 & 0.21 \\
1.425 -- 1.452 & 0.020821 & 6.2e-05 & 0.29 & 0.21 \\
1.452 -- 1.476 & 0.020691 & 5.1e-05 & 0.28 & 0.22 \\
1.476 -- 1.502 & 0.020682 & 6.9e-05 & 0.26 & 0.23 \\
1.502 -- 1.528 & 0.020679 & 6.7e-05 & 0.25 & 0.23 \\
1.528 -- 1.552 & 0.020509 & 6.0e-05 & 0.23 & 0.25 \\
1.552 -- 1.579 & 0.020480 & 6.4e-05 & 0.22 & 0.24 \\
1.579 -- 1.603 & 0.020500 & 5.6e-05 & 0.20 & 0.24 \\
1.603 -- 1.629 & 0.020514 & 6.5e-05 & 0.19 & 0.25 \\
\enddata
\end{deluxetable}

@@

We also performed a ``free" retrieval that varied the abundances of CH$_4$, H$_2$O and NH$_3$ with no assumption of chemical equilibrium. The $3\,\sigma$ upper limit on methane abundance is $1.2\times10^{-3}$, ruling out the the predicted methane content for a composition with solar C/O and $3\,\times$ solar metallicity (the median metallicity from the chemical equilibrium retrieval).  Based on these tests, we conclude that the atmosphere of WASP-107b is depleted in methane relative to expectations for a solar abundance pattern. By contrast, the water abundance ($6\times10^{-6} - 2\times10^{-3}$) is consistent with predictions for solar composition.
